---
title: CNS Data Bureau - Absenteeism Story 
author: "Stella Canino-Quiñones"
date: "2024-08-01"
output: html_document
---


#Information about the columns
# 1.  Days Member Cnt: The total number of days a student was enrolled (in membership) during the school year. This is determined in MSDE’s reporting at the school, county, and state levels.  
#2. Days Attended Cnt: The aggregate number of days (to the nearest half-day) that the student was in attendance during the school year. This is determined in MSDE’s reporting at the school, county, and state levels.

#Fewer than 5 pct: Indicates the attendance rate is less than 5% of membership days. 

Symbols: 
# * – Used to denote suppressed data when the minimum N size is not met (fewer than 10 students) or when top/bottom coding applies.

<=95.0 – Attendance rate less than or equal to 95%. 
<=5.0 – Attendance rate less than or equal to 5% or fewer than 5 days attended.

#Attendance Data Collection: These student-level attendance data are reported by Local Education Agencies (LEAs) to MSDE via an End-of-Year Attendance Data Collection. Each student record includes entry and exit information, the number of days in attendance, and demographic details. MSDE aggregates data for reporting. 

#Students Absent Fewer Than 5 Days:  
#Percentage of students in the school, county, or state who were absent between zero and four days and who were registered to attend for at least 90 days. 
 
#The percent absent fewer than 5 days is Column J - Fewer 5 Pct 
#The numerator for this statistic is Column K - Fewer 5 Cnt 
#The denominator for this statistic is Column N - More 90 Member Cnt 

## Load libraries and establish settings


```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
library(janitor)
library(knitr)      # For nice tables

```

#Load data for 2025 attendance rates
```{r}

md_2025_attendance_data <- read_csv("~/Documents/GitHub/cns-stella-JOUR2026/code/Attendance_2025.csv")
glimpse(md_2025_attendance_data)
```
#Cleaning data
```{r}
md_2025_attendance_data <- md_2025_attendance_data |> clean_names()
glimpse(md_2025_attendance_data)

```

```{r}
md_2025_attendance_data |> select(-academic_year,-lea_number,-school_number, -create_date)
```

```{r}
#change from chr to num
md_2025_attendance_data <- md_2025_attendance_data |> mutate(attend_rate_pct = as.numeric(attend_rate_pct))

```

```{r}
fixing_it <- md_2025_attendance_data |> filter(school_number != "A")
```

```{r}
fixing_it <- fixing_it |>
  # Filter out rows where the column exactly matches "*"
  filter(chronic_absentee_pct != "*") |>
  # Now safe to convert to numeric
  mutate(chronic_absentee_pct = as.numeric(chronic_absentee_pct))
```

#THE MEDIAN ABSENTEEISM RATE IS 23.7%

```{r}
median(fixing_it$chronic_absentee_pct, na.rm = TRUE)
```

```{r}
fixing_it |>
  # 1. Ensure the column is numeric (if you haven't already saved this change)
  # 2. Calculate the median
  summarise(
    median_rate = median(chronic_absentee_pct, na.rm = TRUE)
  )
```
#Mean is 28.38

```{r}
# Calculate the mean
mean(fixing_it$chronic_absentee_pct, na.rm = TRUE)
```

```{r}
# First, calculate the median so we can plot it
median_val <- median(fixing_it$chronic_absentee_pct, na.rm = TRUE)

# Create the histogram
fixing_it |>
  # Ensure we only plot the valid numbers (removes the NAs from the * entries)
  filter(!is.na(chronic_absentee_pct)) |> 
  ggplot(aes(x = chronic_absentee_pct)) +
  
  # 1. The Histogram bars
  # binwidth = 5 means each bar represents a 5% range (e.g. 20-25%)
  # You can change this to 2 or 10 depending on how detailed you want it
  geom_histogram(binwidth = 5, fill = "purple", color = "white") +
  
  # 2. Add a vertical red line for the Median
  geom_vline(xintercept = median_val, color = "pink", linetype = "dashed", size = 1) +
  
  # 3. Add a label for the Median line
  annotate("text", x = median_val + 2, y = 5, label = paste("Median:", median_val), color = "turquoise", hjust = 0) +
  
  # 4. Clean up labels
  labs(
    title = "Distribution of Chronic Absenteeism in MD Schools",
    subtitle = "Red dashed line indicates the median rate",
    x = "Chronic Absenteeism Rate (%)",
    y = "Number of Schools")
```


