---
title: "Stella Canino - CNS Crime (DC and Baltimore Decrease)"
output: html_notebook
---

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
library(janitor)
library(knitr)      # For nice tables

```

```{r}
bmorecrime202425 <- read_csv("~/Documents/GitHub/cns-stella-JOUR2026/code/crimebaltimore2425.csv")
glimpse(bmorecrime202425)
```


```{r}
#Cleaning data
newbaltdata <- bmorecrime202425 |>
  select(CrimeDateTime, Description)
```

```{r}
head(newbaltdata$CrimeDateTime)
```

```{r}
newbaltdata <- newbaltdata |> mutate(
    CrimeDateTime = mdy_hms(CrimeDateTime)
  )
```

```{r}
newbaltdata <- newbaltdata |> mutate(CrimeDate = as.Date(CrimeDateTime),
CrimeHour = hour(CrimeDateTime)                                     )
```

```{r}
Baltimore2024_2025 <- newbaltdata |> select(CrimeDate, Description)
```

*********************
##This is the 2015-2023 data now.

```{r}
baltcrime2015_2023 <- read_csv("~/Documents/GitHub/cns-stella-JOUR2026/code/crimebalt2015_2023.csv")
```

```{r}
baltcrime2015_2023 <- baltcrime2015_2023 |>
  select(CrimeDateTime, Description)
```


```{r}
baltcrime2015_2023 <- baltcrime2015_2023 |> mutate(
    CrimeDateTime = mdy_hms(CrimeDateTime)
  )
```

```{r}
baltcrime2015_2023 <- baltcrime2015_2023 |> mutate(CrimeDate = as.Date(CrimeDateTime),
CrimeHour = hour(CrimeDateTime)                                     )
```

```{r}
baltcrime2015_2023 <- baltcrime2015_2023 |> select(CrimeDate, Description)
```

*******************
#Unite baltimore crime 2016 to 2025

```{r}
Baltimore_Crime <- bind_rows(baltcrime2015_2023, Baltimore2024_2025)
```

```{r}
Baltimore_Crime |> count(Description, sort = TRUE)
```


```{r}
Baltimore_Crime <- Baltimore_Crime |> 
  mutate(Description = case_when(
    Description %in% c("ROBBERY - COMMERCIAL", "ROBBERY - CARJACKING") ~ "ROBBERY",
    TRUE ~ Description
  ))
```


```{r}
Baltimore_Crime %>%
  count(Description, sort = TRUE)
```


*Notes for Graphics for DC and Baltimore Crime Data*

##For the story, tie the crime decrease in Baltimore and DC to the election.
## AKA add that candidates touts for elections that they helped lower crime, but in reality the picture might be completely different. crime has gone down and nobody knows why. 

##graphics
- Baltimore 
- DC

Types:
- Line Graph of types of crime in timeline (year) (maybe per capita)
- Waffle Chart (to show per capita)
- PG (personal reference)

**************************************************
#This is to do per capita rates for baltimore in different years

*Per capita rate for each crime in each year

```{r}
#Summarize total crimes by year (2015-2025)
 Baltimore_2015_2025 |>
  mutate(Year = year(CrimeDate)) |>
  count(Year, name = "Total_Crimes")
```

```{r}
Baltimore_Crime_Filtered <- Baltimore_Crime %>%
  filter(Description %in% c("ROBBERY", "RAPE", "AGG. ASSAULT", "HOMICIDE"))

# Run your count again on the new dataset to verify it worked!
Baltimore_Crime_Filtered %>%
  count(Description, sort = TRUE)
```



```{r}
#Summarize total types of crime by year (2015-2025)
 Baltimore_Crime_Filtered |>
  mutate(Year = year(CrimeDate)) |>
  count(Year, Description, name = "Total_Crimes")
```

```{r}
#2016: 616,226
#2017: 610,481
#2018: 602,443
#2019: 593,490 
#2020: 583,189 
#2021: 576,575 
#2022: 570,663 
#2023: 567,517
#2024: 568,271
```

```{r}
balt_pop <- tibble(
  Year = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025),
  Population = c(622831,615542,610853,603241,594601,583189,576575,570663,567517,568271,568271)
)
# Note: I repeated the 2024 population (568,271) for 2025 as a placeholder. the 2025 census data hasn't come out yet...
#https://fred.stlouisfed.org/series/MDBALT5POP#
```

```{r}
# 2. Calculate Rate for Each Crime Type
# We filter for Year >= 2016 because we don't have population data for 2015 in your list.
 Baltimore_Crime_Filtered |>
  mutate(Year = year(CrimeDate)) |>             # Start at 2015 to match your population data
  filter(!is.na(Description)) |>       # Remove rows with empty descriptions
  count(Year, Description, name = "Count") |> 
  left_join(balt_pop, by = "Year") |>  # Join the population numbers
  mutate(Rate_Per_1k = (Count / Population) * 1000)

```


```{r}
#CREATE CSV FILE!
write_csv(Baltimore_Crime_Filtered, "Baltimore_Crime_Filtered_2015-2025.csv")
```

```{r}
Baltimore_Crime_Filtered |>
count(CrimeDate, name = "Total_Crimes") 
```



```{r}
Baltimore_Crime_Filtered |>
  # 1. Extract the year from the exact date
  mutate(Year = year(ymd(CrimeDate))) |>
  
  # 2. Group all crimes into that specific year and count them
  count(Year, name = "Total_Crimes") |>
  
  # 3. Sort them chronologically
  arrange(Year)


```

```{r}
Baltimore_Crime_Filtered |>
  # 1. Extract the year
  mutate(Year = year(ymd(CrimeDate))) |>
  
  # 2. Keep ONLY the years 2015 and 2025
  filter(Year %in% c(2015, 2025)) |>
  
  # 3. Count the total crimes for those two years
  count(Year, name = "Total_Crimes") |>
  
  # 4. Sort chronologically (so 2015 is first, 2025 is last)
  arrange(Year) |>
  
  # 5. Calculate the percent change between them
  summarize(
    Percent_Change = ((last(Total_Crimes) - first(Total_Crimes)) / 
                       first(Total_Crimes)) * 100
  )
```


